---
title: "Monitoring Eucalyptus Plantation Health Over Two Decades"
subtitle: "Generalized Additive Models for Spectral Index Gap-Filling and Trend Analysis"
author: "Joselyne MPAYIMANA"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: none
    highlight: null
    self_contained: true
    code_folding: hide
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{css, echo=FALSE}
/* Dark theme */
body, html {
  background-color: #0f1923 !important;
  color: #e8eaf0 !important;
  font-family: "Times New Roman", Times, serif !important;
  font-size: 16px;
  line-height: 1.75;
}
h1, h2, h3, h4 { color: #ffffff !important; }
h1.title { font-size: 2rem; font-weight: 800; }
h3.subtitle { color: #a78bfa !important; font-size: 1.1rem; font-weight: 400; }
h2 { border-bottom: 1px solid #2d3a4f; padding-bottom: 0.4rem; }
a { color: #a78bfa !important; }
p, li { color: #c8cdd8; }

#TOC {
  background: #1a2535 !important;
  border: 1px solid #2d3a4f !important;
  border-radius: 8px;
  padding: 1rem;
}
#TOC a { color: #a78bfa !important; }

pre, code {
  background-color: #1a2535 !important;
  color: #c8cdd8 !important;
  border: 1px solid #2d3a4f !important;
  border-radius: 6px;
  font-size: 0.85em;
}
pre { padding: 1rem; overflow-x: auto; }

.code-folding-btn {
  background: #1a2535 !important;
  border: 1px solid #2d3a4f !important;
  color: #a78bfa !important;
}

table {
  background: #1a2535 !important;
  color: #e8eaf0 !important;
  border-collapse: collapse;
  width: 100%;
  margin: 1rem 0;
}
th {
  background: #0f1923 !important;
  color: #a78bfa !important;
  padding: 0.6rem 1rem;
  border-bottom: 1px solid #2d3a4f;
}
td {
  padding: 0.5rem 1rem;
  border-bottom: 1px solid #2d3a4f;
  color: #c8cdd8;
}

.stat-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(180px, 1fr));
  gap: 1rem;
  margin: 1.5rem 0;
}
.stat-card {
  background: #1a2535;
  border: 1px solid #2d3a4f;
  border-radius: 10px;
  padding: 1.2rem 1rem;
  text-align: center;
}
.stat-card .value { font-size: 1.8rem; font-weight: 700; color: #a78bfa; }
.stat-card .label { font-size: 0.82rem; color: #7a8fa6; margin-top: 0.3rem; }

.tag {
  display: inline-block;
  background: rgba(167,139,250,0.12);
  color: #a78bfa;
  border: 1px solid rgba(167,139,250,0.3);
  padding: 0.2rem 0.7rem;
  border-radius: 4px;
  font-size: 0.8rem;
  margin: 0.15rem;
}

.note-box {
  background: #1a2535;
  border-left: 4px solid #a78bfa;
  border-radius: 0 8px 8px 0;
  padding: 1rem 1.5rem;
  margin: 1.2rem 0;
  color: #c8cdd8;
  font-size: 0.93rem;
}

.figure-caption, .caption {
  color: #7a8fa6 !important;
  font-size: 0.85rem;
  text-align: center;
  margin-top: 0.4rem;
}

img {
  max-width: 100%;
  border-radius: 8px;
  margin: 0.5rem 0;
}

.meta {
  color: #7a8fa6;
  font-size: 0.88rem;
  margin-bottom: 1.5rem;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
```

<div class="meta">
Remote Sensing · Time Series Analysis · Forest Health Monitoring · South America &nbsp;|&nbsp;
University of British Columbia &nbsp;|&nbsp; 2024
</div>

<span class="tag">R</span>
<span class="tag">mgcv</span>
<span class="tag">GAM</span>
<span class="tag">Landsat</span>
<span class="tag">NDVI</span>
<span class="tag">NDMI</span>
<span class="tag">EVI</span>
<span class="tag">Time Series</span>
<span class="tag">Gap Filling</span>
<span class="tag">Forest Health</span>
<span class="tag">REML</span>

---

## Overview

Commercial Eucalyptus plantations are a critical component of the forestry sector across South America,
providing timber, pulpwood, and carbon sequestration services. Monitoring their health over time using
satellite imagery is essential for detecting disturbances early, planning harvests, and assessing
long-term stand productivity. However, Landsat time series data are rarely complete: cloud cover,
sensor gaps, and acquisition scheduling leave substantial periods without valid observations.

This project applies Generalized Additive Models (GAMs) to a 20-year Landsat multispectral time
series from a Eucalyptus stand in South America to characterize vegetation dynamics, fill temporal
gaps in spectral index data, and distinguish seasonal variation from long-term trends. The dataset
spans 2000 to 2019 and captures a striking disturbance event around 2002 to 2003 where NDVI
dropped sharply from values near 0.90 to below 0.35, followed by a multi-year recovery trajectory
that GAMs are well-positioned to describe.

<div class="stat-grid">
<div class="stat-card"><div class="value">20</div><div class="label">Years of Landsat observations (2000-2019)</div></div>
<div class="stat-card"><div class="value">189</div><div class="label">Valid NDVI observations out of 395 scenes</div></div>
<div class="stat-card"><div class="value">76.7%</div><div class="label">Deviance explained by best GAM model</div></div>
<div class="stat-card"><div class="value">3</div><div class="label">Spectral indices modelled: NDVI, NDMI, EVI</div></div>
</div>

---

## Step 1: Load Libraries and Data

```{r load-libraries}
library(mgcv)
library(here)
library(fields)
```

```{r load-data}
my_sdata <- read.csv(
  "C:/Users/joselyn7.stu/OneDrive - UBC/MGEM Course/GEM 540/Lab 10/foliage_measurements.csv"
)

cat("Total observations:", nrow(my_sdata), "\n")
cat("Valid NDVI observations:", sum(!is.na(my_sdata$ndvi)), "\n")
cat("Year range:", min(my_sdata$Year, na.rm = TRUE), "to",
    max(my_sdata$Year, na.rm = TRUE), "\n")
```

---

## Step 2: Explore the Raw Spectral Data

The dataset contains surface reflectance in four Landsat bands and three derived spectral indices.
Before fitting any models, it is important to visualize the raw time series to identify data gaps,
outliers, and the general shape of the temporal signal.

### Landsat Band Reflectance Over Time

```{r band-plots, fig.height=7, fig.cap="Figure 1. Raw Landsat band reflectance values (B1-B4) for the Eucalyptus stand across the full study period (2000-2019). Gaps in the time series reflect missing observations due to cloud cover or sensor issues."}
par(mfrow = c(2, 2),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")

plot(B1 ~ Year_Continuous, my_sdata,
     col = "#60a5fa", pch = 16, cex = 0.7,
     main = "Band 1 (Blue)", xlab = "Year", ylab = "Reflectance")

plot(B2 ~ Year_Continuous, my_sdata,
     col = "#4ade80", pch = 16, cex = 0.7,
     main = "Band 2 (Green)", xlab = "Year", ylab = "Reflectance")

plot(B3 ~ Year_Continuous, my_sdata,
     col = "#f87171", pch = 16, cex = 0.7,
     main = "Band 3 (Red)", xlab = "Year", ylab = "Reflectance")

plot(B4 ~ Year_Continuous, my_sdata,
     col = "#a78bfa", pch = 16, cex = 0.7,
     main = "Band 4 (NIR)", xlab = "Year", ylab = "Reflectance")
```

### Spectral Indices Over Time

Spectral indices combine multiple bands to isolate specific vegetation properties. NDVI captures
photosynthetic activity, NDMI captures canopy moisture content, and EVI reduces atmospheric noise
and soil background effects compared to NDVI.

```{r index-plots, fig.cap="Figure 2. NDVI, EVI, and NDMI time series for the Eucalyptus stand (2000-2019). The sharp decline around 2002-2003 reflects a disturbance event followed by gradual canopy recovery. Note the close correspondence between all three indices during and after the disturbance period."}
par(mfrow = c(1, 3),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")

plot(ndvi ~ Year_Continuous, my_sdata,
     col = "#4ade80", pch = 16, cex = 0.7,
     main = "NDVI", xlab = "Year", ylab = "Index Value")

plot(evi ~ Year_Continuous, subset(my_sdata, evi < 3),
     col = "#f59e0b", pch = 16, cex = 0.7,
     main = "EVI", xlab = "Year", ylab = "Index Value")

plot(ndmi ~ Year_Continuous, my_sdata,
     col = "#60a5fa", pch = 16, cex = 0.7,
     main = "NDMI", xlab = "Year", ylab = "Index Value")
```

<div class="note-box">
The NDVI drop around 2002 to 2003 is striking. Values fall from a pre-disturbance mean of approximately
0.88 to a minimum below 0.35, then recover steadily over the following five to six years. This pattern
is consistent with a harvest or major disturbance event followed by replanting and canopy closure.
Modelling this recovery trajectory is one of the core objectives of the GAM analysis.
</div>

---

## Step 3: Baseline Smoothers

Before fitting formal GAMs, two simpler smoothing approaches are applied to establish a baseline
and illustrate the gap-filling concept.

### LOESS Smoother

```{r loess, fig.cap="Figure 3. LOESS-smoothed NDVI predictions (2003-2020). LOESS fits a locally weighted polynomial and provides a simple non-parametric smooth, but does not separate trend from seasonality."}
reg_loess <- loess(ndvi ~ Year_Continuous, my_sdata)

new_data              <- expand.grid(Year = 2003:2020, doy = 1:365)
new_data$Year_Continuous <- new_data$Year + new_data$doy / 365
pred_loess            <- predict(reg_loess, newdata = new_data)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")
plot(new_data$Year_Continuous, pred_loess,
     type = "l", col = "#4ade80", lwd = 1.5,
     xlab = "Year", ylab = "Predicted NDVI",
     main = "LOESS Smoothed NDVI Predictions")
```

### Thin Plate Spline

```{r tps, fig.cap="Figure 4. Thin Plate Spline (TPS) fitted to the NDVI time series. The TPS minimizes a roughness penalty globally, providing a smooth interpolation across the full time range."}
reg_tps <- with(my_sdata, Tps(x = Year_Continuous, Y = ndvi))

new_data  <- with(new_data, new_data[order(Year_Continuous), ])
pred_tps  <- predict(reg_tps, new_data$Year_Continuous)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")
plot(new_data$Year_Continuous, pred_tps,
     type = "l", col = "#a78bfa", lwd = 1.5,
     xlab = "Year", ylab = "Predicted NDVI",
     main = "Thin Plate Spline Fitted NDVI")
points(ndvi ~ Year_Continuous, my_sdata, cex = 0.3, col = "#c8cdd8")
```

---

## Step 4: GAM — Temporal Trend

The first formal GAM models only the long-term temporal trend in NDVI using a Gaussian process smooth
on the continuous year variable. This captures the overall shape of the disturbance and recovery without
accounting for within-year seasonality.

```{r gam-01, fig.cap="Figure 5. Smooth effect of time on NDVI from the temporal-trend-only GAM (reg_01b). The Gaussian process basis function is well-suited for this type of temporal autocorrelation structure."}
reg_01b <- gam(ndvi ~ s(Year_Continuous, bs = "gp"),
               data   = my_sdata,
               method = "REML")

cat("=== MODEL: reg_01b (Temporal Trend Only) ===\n")
summary(reg_01b)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")
plot(reg_01b,
     col = "#a78bfa", shade = TRUE, shade.col = "#a78bfa33",
     main = "Smooth Effect: Year on NDVI")
```

### Predictions with Confidence Intervals

```{r gam-01-pred, fig.cap="Figure 6. GAM predictions with 95% confidence intervals for the temporal-trend-only model. The grey band widens in the post-disturbance recovery period, reflecting greater uncertainty in the model estimates during that phase."}
preds <- predict(reg_01b,
                 newdata = data.frame(Year_Continuous = seq(2000, 2020, 0.1)),
                 se.fit  = TRUE)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")

plot(seq(2000, 2020, 0.1), preds$fit,
     type = "n", ylim = c(0.3, 1.0),
     xlab = "Year", ylab = "NDVI",
     main = "GAM Temporal Trend with Confidence Interval")

polygon(x = c(seq(2000, 2020, 0.1), rev(seq(2000, 2020, 0.1))),
        y = c(preds$fit + preds$se.fit,
              rev(preds$fit - preds$se.fit)),
        col = "#a78bfa33", border = "#a78bfa66")

lines(seq(2000, 2020, 0.1), preds$fit, col = "#a78bfa", lwd = 2)
points(ndvi ~ Year_Continuous, my_sdata, cex = 0.4, col = "#c8cdd8")
```

### Model Diagnostics

```{r gam-01-diag, fig.height=6, fig.cap="Figure 7. GAM diagnostic plots for the temporal-trend-only model. These check residual normality, homoscedasticity, and the adequacy of basis dimension k."}
par(mfrow = c(2, 2),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")
gam.check(reg_01b)
```

---

## Step 5: GAM — Seasonal Component

The second GAM isolates the within-year seasonal cycle using a cyclic cubic spline on day-of-year.
It is fitted only to post-2008 data, when the stand has fully recovered from the disturbance and
exhibits a stable seasonal signal.

```{r gam-02, fig.cap="Figure 8. Smooth seasonal effect (day-of-year) on NDVI from the seasonal-only GAM (reg_02). The cyclic constraint forces the curve to be equal at day 1 and day 365, which is ecologically appropriate for annual cycles."}
reg_02 <- gam(ndvi ~ s(doy, bs = "cc"),
              data   = subset(my_sdata, Year > 2008),
              method = "REML")

cat("=== MODEL: reg_02 (Seasonal Component Only) ===\n")
summary(reg_02)

par(mfrow = c(1, 1),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")

plot(reg_02,
     shade     = TRUE,
     shade.col = "#4ade8033",
     scale     = 0,
     col       = "#4ade80",
     main      = "Seasonal Effect on NDVI (Post-2008)")
```

### Seasonal Prediction

```{r gam-02-pred, fig.cap="Figure 9. GAM seasonal predictions across a full calendar year (days 1-365) with observed post-2008 data overlaid. Peak NDVI occurs in late summer, consistent with peak growing season for Eucalyptus."}
pred_02 <- predict(reg_02,
                   newdata = data.frame(doy = 1:365),
                   se.fit  = TRUE)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")

plot(1:365, pred_02$fit,
     type = "n", ylim = c(0.6, 1.0),
     xlab = "Day of Year", ylab = "NDVI",
     main = "Seasonal GAM Prediction with Confidence Interval")

polygon(x = c(seq(1, 365), rev(seq(1, 365))),
        y = c(pred_02$fit + pred_02$se.fit,
              rev(pred_02$fit - pred_02$se.fit)),
        col = "#4ade8033", border = "#4ade8066")

lines(1:365, pred_02$fit, col = "#4ade80", lwd = 2)
points(ndvi ~ doy, subset(my_sdata, Year > 2008),
       cex = 0.5, col = "#c8cdd8")
```

---

## Step 6: GAM — Combined Trend and Seasonality (Best Model)

The third and best-performing model combines the long-term temporal trend with the within-year
seasonal cycle as separate additive smooth components. This is the key step for gap-filling because
it allows predictions to be ecologically informed: a gap during a high-growth summer period will
be filled with a higher predicted value than one during a dormant winter period.

```{r gam-03, fig.cap="Figure 10. Smooth components of the combined trend-plus-seasonality GAM (reg_03). Left: long-term temporal trend. Right: within-year seasonal cycle. Both are plotted on their own scale."}
reg_03 <- gam(ndvi ~ s(Year, bs = "gp") + s(doy, bs = "cc", k = 20),
              data   = subset(my_sdata, Year >= 2004),
              method = "REML")

cat("=== MODEL: reg_03 (Trend + Seasonality) ===\n")
summary(reg_03)

par(mfrow = c(1, 2),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")

plot(reg_03,
     shade     = TRUE,
     shade.col = "#a78bfa33",
     scale     = 0,
     col       = "#a78bfa")
```

<div class="note-box">
The combined model explains 76.7% of the deviance in NDVI with an adjusted R² of 0.782, a substantial
improvement over the trend-only model. The Gaussian process smooth on year captures the post-disturbance
recovery trajectory, while the cyclic cubic spline on day-of-year captures the seasonal pulse of
photosynthetic activity.
</div>

### Gap-Filling: Full Time Series Prediction

```{r gam-03-pred, fig.cap="Figure 11. Full gap-filled NDVI time series (2003-2020) from the combined trend-plus-seasonality GAM. The grey band shows the 95% confidence interval. The model successfully reconstructs the disturbance and recovery trajectory while capturing seasonal variation throughout."}
newdata              <- expand.grid(doy = 1:365, Year = 2003:2020)
newdata$Year_Continuous <- with(newdata, Year + doy / 365)
pred_03              <- predict(reg_03, newdata = newdata, se.fit = TRUE)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")

plot(ndvi ~ Year_Continuous, my_sdata,
     xlim = c(2003, 2020), type = "n",
     xlab = "Year", ylab = "NDVI",
     main = "Gap-Filled NDVI: Combined Trend and Seasonality Model")

with(newdata,
     polygon(x = c(Year_Continuous, rev(Year_Continuous)),
             y = c(pred_03$fit + pred_03$se.fit,
                   rev(pred_03$fit - pred_03$se.fit)),
             col = "#a78bfa22", border = "#a78bfa44"))

lines(pred_03$fit ~ newdata$Year_Continuous,
      col = "#a78bfa", lwd = 1.5)
points(ndvi ~ Year_Continuous, my_sdata,
       cex = 0.5, col = "#c8cdd8", pch = 16)
```

---

## Step 7: Applying the Model to NDMI

The same combined trend-plus-seasonality model is applied to NDMI to assess whether moisture
dynamics follow a similar pattern. NDMI is sensitive to canopy water content and often provides
a cleaner signal than NDVI in dense Eucalyptus canopies.

```{r gam-03b, fig.cap="Figure 12. Smooth components of the combined GAM for NDMI (reg_03b). NDMI models generally show better fit than NDVI due to lower high-frequency noise in moisture-sensitive spectral indices."}
reg_03b <- gam(ndmi ~ s(Year, bs = "gp") + s(doy, bs = "cc", k = 20),
               data   = subset(my_sdata, Year >= 2004),
               method = "REML")

cat("=== MODEL: reg_03b (NDMI Trend + Seasonality) ===\n")
summary(reg_03b)

par(mfrow = c(1, 2),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")

plot(reg_03b,
     shade     = TRUE,
     shade.col = "#60a5fa33",
     scale     = 0,
     col       = "#60a5fa")
```

### NDMI Model Diagnostics

```{r gam-03b-diag, fig.height=6, fig.cap="Figure 13. Diagnostic plots for the NDMI combined model. Residuals show good normality and homoscedasticity, confirming model adequacy."}
par(mfrow = c(2, 2),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")
gam.check(reg_03b)
```

### NDMI Gap-Filled Predictions

```{r gam-03b-pred, fig.cap="Figure 14. Gap-filled NDMI predictions (2003-2020) from the combined trend-plus-seasonality model. The tighter confidence intervals compared to NDVI reflect the lower noise level in the moisture index."}
pred_03b <- predict(reg_03b, newdata = newdata, se.fit = TRUE)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")

plot(ndmi ~ Year_Continuous, my_sdata,
     xlim = c(2003, 2020), type = "n",
     xlab = "Year", ylab = "NDMI",
     main = "Gap-Filled NDMI: Combined Trend and Seasonality Model")

with(newdata,
     polygon(x = c(Year_Continuous, rev(Year_Continuous)),
             y = c(pred_03b$fit + pred_03b$se.fit,
                   rev(pred_03b$fit - pred_03b$se.fit)),
             col = "#60a5fa22", border = "#60a5fa44"))

lines(pred_03b$fit ~ newdata$Year_Continuous,
      col = "#60a5fa", lwd = 1.5)
points(ndmi ~ Year_Continuous, my_sdata,
       cex = 0.5, col = "#c8cdd8", pch = 16)
```

---

## Step 8: Adding a Trend-Season Interaction

The fourth model tests whether the amplitude of seasonal variation changed over the study period
by adding a tensor product interaction term between year and day-of-year. This is a more
ecologically complex question: did the Eucalyptus stand's seasonal response to climate shift
as the canopy recovered from disturbance?

```{r gam-04, fig.cap="Figure 15. Smooth components of the interaction model for NDMI (reg_04), including the tensor product interaction term (ti). The significant interaction confirms that seasonal moisture amplitude shifted over the study period."}
reg_04 <- gam(ndmi ~ s(Year) + s(doy, bs = "cc") + ti(doy, Year),
              data   = my_sdata,
              method = "REML",
              select = TRUE)

cat("=== MODEL: reg_04 (NDMI Trend + Seasonality + Interaction) ===\n")
summary(reg_04)

par(mfrow = c(1, 3),
    bg    = "#0f1923",
    col.axis = "#7a8fa6",
    col.lab  = "#c8cdd8",
    col.main = "#ffffff",
    fg       = "#2d3a4f")

plot(reg_04,
     shade     = TRUE,
     shade.col = "#60a5fa33",
     scale     = 0,
     col       = "#60a5fa")
```

### Comparing Interaction vs Additive Models

```{r model-comparison, fig.cap="Figure 16. Comparison of predictions from the additive (reg_03b, red) and interaction (reg_04, blue) NDMI models (2004-2020). The interaction model allows the seasonal amplitude to shift across years while the additive model assumes it remains constant."}
pred_04 <- predict(reg_04, newdata = newdata, se.fit = TRUE)

par(bg = "#0f1923", col.axis = "#7a8fa6",
    col.lab = "#c8cdd8", col.main = "#ffffff", fg = "#2d3a4f")

plot(ndmi ~ Year_Continuous, my_sdata,
     xlim = c(2004, 2020), type = "n",
     xlab = "Year", ylab = "NDMI",
     main = "Model Comparison: Additive vs Interaction NDMI Models")

lines(pred_04$fit ~ newdata$Year_Continuous,  col = "#60a5fa", lwd = 1.5)
lines(pred_03b$fit ~ newdata$Year_Continuous, col = "#f87171", lwd = 1.5)
points(ndmi ~ Year_Continuous, my_sdata, cex = 0.4, col = "#c8cdd8", pch = 16)

legend("bottomright",
       legend = c("Interaction model (reg_04)", "Additive model (reg_03b)"),
       col    = c("#60a5fa", "#f87171"),
       lwd    = 2,
       bg     = "#1a2535",
       text.col = "#c8cdd8",
       border = "#2d3a4f")
```

---

## Model Performance Summary

```{r summary-table}
results <- data.frame(
  Model    = c("reg_01b", "reg_02", "reg_03", "reg_03b", "reg_04"),
  Index    = c("NDVI", "NDVI", "NDVI", "NDMI", "NDMI"),
  Components = c(
    "Temporal trend only",
    "Seasonal cycle only (post-2008)",
    "Trend + seasonality",
    "Trend + seasonality",
    "Trend + seasonality + interaction"
  ),
  Deviance_Explained = c("~55%", "~40%", "76.7%", ">78%", "Significant interaction"),
  Adj_R2             = c("~0.54", "~0.38", "0.782", ">0.78", "N/A")
)

knitr::kable(results,
             col.names = c("Model", "Index", "Components",
                           "Deviance Explained", "Adj. R2"),
             caption   = "Table 1. Summary of GAM model performance across all fitted models.")
```

---

## Key Findings and Reflections

The combined trend-plus-seasonality model (reg_03) explained 76.7% of the deviance in NDVI with
an adjusted R2 of 0.782, substantially outperforming the trend-only model. The Gaussian process
smooth captured the sharp decline and multi-year recovery following the 2003 disturbance, while
the cyclic cubic spline captured the within-year seasonal pulse of photosynthetic activity.

The NDMI models showed a better fit than NDVI, likely because moisture-sensitive spectral indices
exhibit less high-frequency noise in densely vegetated Eucalyptus canopies. The tensor interaction
model for NDMI confirmed a statistically significant interaction between year and day-of-year,
meaning the amplitude of seasonal moisture variation shifted over the course of the study period.

The GAMs here appear to slightly underfit in periods of high seasonal variation, where confidence
intervals widen. This is a deliberate trade-off: the smoothing penalty prevents overfitting to
noise while accepting some bias in high-variability periods. To reduce overfitting if it were a
concern, increasing the smoothing penalty or reducing the number of knots `k` in the spline would
constrain the curve further.

One important limitation of these models is that they are entirely aspatial. They describe temporal
dynamics at a single stand without accounting for spatial autocorrelation among neighboring stands.
Two natural extensions would be spatial lag or error models, which explicitly incorporate a spatial
weights matrix, and Geographically Weighted Regression, which allows model coefficients to vary
across the landscape.

A parametric alternative to GAMs would be a linear mixed-effects model with polynomial terms for
year and day-of-year. These are more interpretable since each coefficient directly describes the
direction and magnitude of a predictor's effect. A non-parametric alternative such as Random Forest
can capture more complex nonlinear relationships, but understanding how a variable drives predictions
requires additional tools such as SHAP values rather than simply reading off model coefficients.

---

## References

Wellington, M.J., Lawes, R., and Kuhnert, P. (2023). A framework for modelling spatio-temporal
trends in crop production using generalised additive models. *Computers and Electronics in
Agriculture*, 212, 108111.

Kinane, S.M., et al. (2021). A model to estimate leaf area index in loblolly pine plantations
using Landsat 5 and 7 images. *Remote Sensing*, 13(6), 1140.

---

<div style="color:#7a8fa6; font-size:0.82rem; text-align:center; margin-top:2rem;">
Joselyne MPAYIMANA &nbsp;|&nbsp; Master of Geomatics for Environmental Management, UBC &nbsp;|&nbsp; 2024
</div>
